---
title: "Stat5405 Homework 4 Open Question"
author: "Giovanni Lunetta"
date: "September 27, 2023"
format: html
embed-resources: true
theme: cosmo
code-line-numbers: true
number_examples: true
number_section: true
number_chapters: true
linkcolor: blue
editor: visual
fig-cap-location: top
---

**Q4. Open question. Not graded for credit.**
Choose an application domain that interests you. Find/download an interesting data set in this domain which lends to analysis using one-factor or two-factor ANOVA analysis. Carry out a comprehensive analysis (testing assumptions, testing a suitable hypothesis, estimating effects etc.). Summarize as a report. You can share this on Discussion Board, and include on your Github page. 

Overview:

The National Basketball Association (NBA) is a men’s professional basketball league in North America. The league is composed of 30 teams (29 in the United States and 1 in Canada) and is one of the four major professional sports leagues in the United States and Canada. The NBA is an active member of USA Basketball (USAB), which is recognized by FIBA (also known as the International Basketball Federation) as the national governing body for basketball in the United States. The NBA is one of the four major professional sports leagues in the United States and Canada. NBA players are the world’s best paid athletes by average annual salary per player.

In this analysis, we are interested in exploring the relationship between team pace and the division that a team belongs to. Team pace is a basketball statistic that estimates the number of possessions per 48 minutes by a team. A possession is defined as the time a team has the ball until it scores, misses a shot, commits a turnover, or commits a foul that is not a shooting foul. The division that a team belongs to is a categorical variable that indicates the division that a team belongs to. There are five divisions in the NBA: Atlantic, Central, Southeast, Northwest, Pacific, and Southwest.

```{r}
library(dplyr)
library(tidyverse)
library(car)

# Load the modified dataset with the 'Division' column
data <- read.csv("/Users/giovanni-lunetta/uconn_masters/stat5405/hw_4/merged_with_mapped_salaries_and_conference_df.csv")
```

Null Hypothesis: The mean team pace is the same for all divisions.

$H_0$: $\mu_1 = \mu_2 = \mu_3 = \mu_4 = \mu_5$

where $\mu_i$ is the mean team pace for division $i$.

Alternative Hypothesis: The mean team pace is different for at least one division.

$H_a$: $\mu_i \neq \mu_j$ for at least one pair of divisions $i$ and $j$.

Assumptions (WILL BE VALIDATED AFTER RUNNING ANOVA MODEL):

Normality:

We assume that the team pace for each division is normally distributed.

Independence:

We assume that the team pace for each division is independent of the team pace for the other divisions.

Equal Variance:

```{r}
# Create boxplots by division using ggplot
ggplot(data, aes(x=as.factor(Division), y=Team_Pace, fill=as.factor(Division))) + 
    geom_boxplot(show.legend = FALSE) +
    labs(x="Division", y="Team Pace") +
    theme_minimal() +
    ggtitle("Boxplots of Team Pace by Division")
```
We assume that the variance of the team pace for each division is equal.

```{r}
# Convert Division and Conference to factors
data$Division <- as.factor(data$Division)
data$Conference <- as.factor(data$Conference)
```

```{r}
# Run one-way ANOVA
anova_result <- aov(Team_Pace ~ Division, data = data)

# View the summary of the ANOVA result
summary(anova_result)
```

Interpretation:

The p-value is the probability that you obtain the observed results, or more extreme, given that the null hypothesis is true. A smaller p-value (< 0.05) usually indicates that you can reject the null hypothesis.

In this case, the p-value obtained from the one-way ANOVA is 0.233. Since this value is greater than the common alpha level of 0.05, we fail to reject the null hypothesis.

Conclusion:

Based on our analysis, there is not sufficient evidence to conclude that there is a significant difference in the mean team pace among the different divisions, at the 0.05 significance level. In other words, we do not have enough statistical evidence to say that the mean team pace is different for at least one division.

Further Remarks:

Even though there isn’t a significant difference in the mean team pace among the divisions, it could be insightful to explore other factors or variables that might affect team pace. Additionally, exploring the interactions between different variables might reveal significant effects that are not apparent when considering each variable separately.

Validating Assumptions:

Since the p-value is not less than 0.05, strictly speaking, there is no need to check the assumptions of normality and equal variance (homogeneity of variance), as we do not proceed with post hoc tests when you fail to reject the null hypothesis in ANOVA.

Next Steps:

What if there is a difference in pace between conferences (Eastern and Western)?

Null Hypothesis: The mean team pace is the same for both conferences.

$H_0$: $\mu_{Eastern} = \mu_{Western}$

where $\mu_{Eastern}$ is the mean team pace for the Eastern conference and $\mu_{Western}$ is the mean team pace for the Western conference.

Alternative Hypothesis: The mean team pace is different for the two conferences.

$H_a$: $\mu_{Eastern} \neq \mu_{Western}$

Assumptions (WILL BE VALIDATED AFTER RUNNING ANOVA MODEL):

Normality:

We assume that the team pace for each conference is normally distributed.

Independence:

We assume that the team pace for each conference is independent of the team pace for the other conference.

Equal Variance:

```{r}
# Create boxplots by conference using ggplot
ggplot(data, aes(x=as.factor(Conference), y=Team_Pace, fill=as.factor(Conference))) + 
    geom_boxplot(show.legend = FALSE) +
    labs(x="Conference", y="Team Pace") +
    theme_minimal() +
    ggtitle("Boxplots of Team Pace by Conference")
```
We assume that the variance of the team pace for each conference is equal.

```{r}
# Run one-way ANOVA for Conference
anova_result_conference <- aov(Team_Pace ~ Conference, data = data)

# View the summary of the ANOVA result
summary(anova_result_conference)
```

Interpretation:

The p-value is the probability that you obtain the observed results, or more extreme, given that the null hypothesis is true. A smaller p-value (< 0.05) usually indicates that you can reject the null hypothesis.

In this case, the p-value obtained from the one-way ANOVA is 0.0383. Since this value is less than the common alpha level of 0.05, we reject the null hypothesis.

Conclusion:

Based on our analysis, there is sufficient evidence to conclude that there is a significant difference in the mean team pace between the two conferences, at the 0.05 significance level. In other words, we have enough statistical evidence to say that the mean team pace is different between the conferences.

Validating Assumptions:

Normality:

Null Hypothesis: The residuals $e_{ij}$ are normally distributed.

$H_0$: $e_{ij} \sim N(0, \sigma^2)$

where $e_{ij}$ is the residual for the $j$th observation in the $i$th conference.

Alternative Hypothesis: The residuals $e_{ij}$ are not normally distributed.

$H_a$: $e_{ij} \nsim N(0, \sigma^2)$

```{r}
aovfits <- fitted(anova_result_conference)
aovres <- residuals(anova_result_conference)
car::qqPlot(aovres, main = "Q-Q Plot of Residuals", pch = 19, col = 2, cex = 0.7)
```
```{r}
shapiro.test(aovres)
```
The $p$-value of 0.1786 from the Shapiro-Wilk significance test for normality of $e_{ij}$ validates the normality assumption.

Equal Variance:

Null Hypothesis: The variance of the team pace for each conference is equal.

$H_0$: $\sigma_{Eastern}^2 = \sigma_{Western}^2$

where $\sigma_{Eastern}^2$ is the variance of the team pace for the Eastern conference and $\sigma_{Western}^2$ is the variance of the team pace for the Western conference.

Alternative Hypothesis: The variance of the team pace for each conference is not equal.

$H_a$: $\sigma_{Eastern}^2 \neq \sigma_{Western}^2$

```{r}
library(lawstat)

levene_result_lawstat <- levene.test(data$Team_Pace, factor(data$Conference))

print(levene_result_lawstat)
```
The $p$-value of 0.9101 from the Levene significance test for equal variance validates the equal variance assumption.

Overall Conclusion:

Based on our analysis, there is sufficient evidence to conclude that there is a significant difference in the mean team pace between the two conferences, at the 0.05 significance level. In other words, we have enough statistical evidence to say that the mean team pace is different between the conferences.